Use of communication channels in 2x2 strategic games

### 2025-07-16 manuscript/manuscript.Rmd
fix typos

---
title: "Use of communication channels in 2x2 strategic games"
authors:
  - name: Veronica Boyce
    department: Department of Psychology
    affiliation: Stanford University
    email: vboyce@stanford.edu
  - name: Avidit Acharya
    department: Department of Political Science
    affiliation: Stanford University
  - name: Michael C. Frank
    department: Department of Psychology
    affiliation: Stanford University
abstract: |
  How do we use language to coordinate with each other? This question is of interest both from the cognitive science of language perspective as well as the game theory perspective. Here we attempt to bridge the gap between the empirical tradition of studying language use for coordination in reference games with the theoretical and analytic framework of game theory. We conducted a series of 5 real-time online experiments where pairs of participants were matched to play a sequence of 40 2x2 games together (total N=616). Across the experiments, we varied the exact structure of trials, but all experiments involved trials with reward structures modeled on Prisoner's Dilemma (PD) and Bach or Stravinsky (BoS). We manipulated the presence or absence of a chat interface as a between-pairs manipulation. Despite low levels of use of the communication channel, we found slight differences in outcome between the chat and nochat groups, where pairs who could communicate were slightly more likely to reach coordinated beneficial outcomes in BoS games. In contrast to the predictions of rational actors, participants in both chat and no-chat conditions generally aligned on both cooperating rather than both defecting in PD. Overall, we were limited in our ability to discern the role of communication channels in game strategies due to low uptake in using the communication channels when available.

@@

In studies of conversation, there's work on how people align with each other in their word use when doing joint task and how repeatedly referring to the same targets in a matching game leads to the formation of linguistic conventions. As reference is a key part of other linguistic speech acts such as negotiation, coordinating actions, or persuasion, one might wonder whether referential coordination can also be seen during interactions in pursuit of these more complex goals. In a similar vein, more complex situations may allow for linguistic alignment and convention formation not just of terms of reference but of terms for entire strategic approaches. One could also imagine alignment in approach, creating a conventional structure for how to coordinate or negotiate. (todo possible citations)

@@

In prior work, we tried to look at the role of language in a 3-player negotiation game where incentives were either fully-aligned or competitive between players [@mankewitz2021; @boyce2023b]. We found some evidence that language use improved performance, and that the referential language used to pick out targets showed some signs of conventionalization over time. However, this was a game we had created, so there were not theoretical or analytic predictions about what the optimal strategies were either with or without language, and therefore, when or how communication (via language) might improve performance.

Game theory provides a formalism for understanding rational action when different actors have different incentive structures over outcomes that are jointly determined by the choices of all the actors. There is some theoretical work analyzing how communication availability may shift the strategies in games and thus when communicative acts are credible signals versus cheap talk [@cooper1992; @foerster2021; @baliga2002]. Game theory economic games have been used to empirically study questions such as social prediction [@vanbaar2021], emotional signals [@demelo2020], and coordination [@duguid2014]. Some studies test predictions about the role of communication in economic games [@crandall2018; @cross; @crawford2019a]. (todo: check/reread citations, possible find more citations?)

@@

In Bach or Stravinsky (Figure \ref{payoff-BoS}), both players only receive rewards if they choose the same option. However, player X prefers the AA option and B prefers the BB option. Thus, there are two Nash equilibria, either AA or BB. In a one-shot game of BoS, there is no way for participants to break the symmetry. Communication is predicted to be potentially useful, if it allows participants to break this symmetry and agree on which cell to align on.

@@

## Participants
The number of participants included in each condition (after incomplete games were excluded) is shown in Table \ref{participants}. We recruited participants via Prolific and directed them to our website where they were paired up to play a game with another participant in real time. 
## Procedure

All experiments were implemented in Empirica v1 [@almaatouq2021]. The experiment consisted first of consent and instruction.

@@

After the three minutes, participants moved onto the main experiment.  On each trial, they saw a payout grid for differently colored treasure chests and were asked to click which chest they wanted to open (Figure \ref{ui}). Each pair completed 40 trials, with the payoffs dependent on the experiment condition (see below). Participants in the "chat" condition could message their partner freely during the trials using a chat interface; participants in the "no chat" condition did not have access to the chat during the main experiment.

In experiments 1,2,3 and 5, participants saw a running bonus score for themselves and their partner, as their bonus payment was proportional to the sum of the rewards they earned. On experiment 4, participants were instead paid proportional to the results of 4 randomly selected trials, and were not shown a cumulative payoff score.

@@

## Trial types 
We sampled different types of trials across different experiments. For each trial type, we constructed trials by randomly sampling reward values subject to some constraints. Examples of each trial type are shown in Figure \ref{payoff-matrix}

@@

**Expt 4** (https://osf.io/fymjc): In experiment 4, we aimed for a set-up that was closer to the predictions from game theory.We removed the points counter, given that participants in prior experiments sometimes used the visible money earned so far as a locus for coordination. We switched from providing a bonus proportional to the overall points earned to an incentive compatible bonus that sampled 4 trials at random and paid out those points. Because the games that mattered were unknown and randomly sampled, we reasoned that the rational strategy might be closer to the formal predictions of one-shot games. Trials were sampled from a 50/50 distribution of easy PD and normal BoS.

@@

exit <- exit_1 |>
  bind_rows(exit_2) |>
  bind_rows(exit_3) |>
  bind_rows(exit_4) |>
  bind_rows(exit_5) |>
  select(expt, chat_cond, pct) |>
  mutate(chat_cond = ifelse(chat_cond == "chat", "chat", "nochat"))

@@

```{r bosquad, fig.cap="For BoS trials, how frequently pairs made coordinated selections that recieved rewards. \\label{bosquad} ", fig.height=3, fig.width=8}
outcome %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond), y = outcome_parity, color=combo_type, group=combo_type)) +
  facet_grid(~ expt) +
  scale_color_manual(values=pal)+
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  coord_cartesian(ylim=c(0,1))+
  labs(y = "Good outcome?", x="Chat condition", color="")+
  theme(legend.position="bottom")
```
For PD trials, there are three distinct outcomes. If both players cooperate, the outcome is AA and each gets a moderately high payoff. If both players defect, the outcome is BB, the Nash equilibrium, and each player get a low payoff. If one player cooperates and the other defects, the outcome is AB (or BA), and one player get a high payoff and the other gets 0. From a perspective of wefare-maximization, the selection with best total reward is AA in easy PD and AB/BA in hard PD. We expect normal PD to generally be the of the type easy PD; and again note that the errors in displaying off-diagonal rewards would primarily affect hard PD type trials, where off diagonal options are more attractive. The rates of selecting each of the three outcome options by experiment and conditions are shown in Figure \ref{pdquad}.

@@

Overall, the results based on condition assignment are consistent with the prediction that language would be more beneficial for BoS than for PD. However, condition differences were small and often noisy (presumably due to low update of using language), and so the experiments do not provide compelling evidence. Additionally, the analytic import of using well-studied game theory scenarios with predictions about the utility of language is weakened by the inconsistency between predicted behaviors in the PD condition and actual selections.

@@

```{r}
outcome_chatted <- outcome |> 
  mutate(chatted = case_when(
    chat_cond =="nochat" ~ "nochat",
    is_chat == 1 ~ "chatty",
    is_chat == 0 ~ "quiet"
  ) |> factor(levels=c("chatty", "quiet", "nochat")))
```

We repeat the quadrant analyses, now with a three-way split between "chatty" games who used the chat on the given trial, "quiet" games who were in the chat condition, but produced no chat messages on the given trial, and "nochat" games were assigned to the no chat condition.

```{r, fig.cap="For BoS trials, how frequently pairs made coordinated selections that recieved rewards as a function of whether chat was available or used that trial. \\label{bosquad-uptake} ", fig.height=3, fig.width=8}
outcome_chatted %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = chatted, y = outcome_parity, color=combo_type, group=combo_type)) +
  facet_grid(~ expt) +
    coord_cartesian(ylim=c(0,1))+
  scale_color_manual(values=pal)+
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  labs(y = "Good outcome?", x="Chat usage on trial", color="")+
  theme(legend.position="bottom")
```
For BoS results, shown in Figure \ref{bosquad-uptake}, chatty trials had the best outcomes, with high levels of aligning. Depending on experiment, quiet trials generally patterned with the nochat condition, expect in the first experiment that was entirely BoS. Thus, it appears that there may be slight benefits to chatting that spill over onto other trials, perhaps via setting up conventions for determining what strategy to use to break the symmetry. Strategy set-up may be easiest when the trials are most homogeneous. However, most of the overall slight benefits associated with the chat condition are being driven by the small number of games and trials that used the chat.

@@

We similarly break down the PD results in terms of language use per trial in Figure \ref{pdquad}. Trials that used chat were slightly more consistent at selecting the cooperate option on easy and normal PD (although this in particular could reflect selection effects for careful or attentive participants). The larger difference in selections between trials that used chat and those that did not was on the hard PD trials where selection of the AB/BA option appears to be primarily driven by chat-usage.

Overall, the slight differences observed in the chat condition and nochat conditions appear to primarily come from the relatively small number of trials that actually used the chat while quiet trials generally patterned with behavior from the nochat condition.

@@

Despite low levels of use of the communication channel, we found slight differences in outcome between the chat and nochat groups, where pairs who could communicate were slightly more likely to reach coordinated beneficial outcomes in BoS games and slightly more likely to choose welfare-maximizing unequal outcomes in hard PD games. In contrast to the predictions of rational actors, participants in both chat and no-chat conditions generally aligned on both cooperating rather than both defecting in PD. The tendency to cooperate has previously been observed in experiments with repeated PD-style games [@rand2013]. Overall, we were limited in our ability to discern the role of communication channels due to low uptake in using the communication channels when available.

@@

In conclusion, we still think there are interesting questions about how communication, and linguistic communication in particular, might affect what strategies around coordination and competition can be used in strategic games. Coming from the perspective of convention formation in reference games, we are still interested in when and how people can form procedural or "meta" conventions around how to negotiate with each other rather than just around referents. However, over the course of these experiments, we have learned that there are methodological and implementational barriers. We were able to mostly overcome initial issue with participants thinking they were playing with bots by adding an initial short free chat between each partnership. However, we were not able to overcome other problems. We had difficulties finding tasks that have tractable theoretical predictions but are also sufficiently engaging that humans will take them seriously, and compromised on a design that was not ideal from either perspective. Additionally, while we previously had no problems getting participants to use linguistic communication in online chat-box mediated tasks where the goal was reference [@boyce2024], we had difficulty getting participants to use chat in strategic contexts [both here and in @mankewitz2021].

>>>>

---
title: "Use of communication channels in 2x2 strategic games"
authors:
  - name: Veronica Boyce
    department: Department of Psychology
    affiliation: Stanford University
    email: vboyce@stanford.edu
  - name: Avidit Acharya
    department: Department of Political Science
    affiliation: Stanford University
  - name: Michael C. Frank
    department: Department of Psychology
    affiliation: Stanford University
abstract: |
  How do we use language to coordinate with each other? This question is of interest both from the cognitive science of language perspective as well as the game theory perspective. Here we attempt to bridge the gap between the empirical tradition of studying language use for coordination in reference games with the theoretical and analytic framework of game theory. We conducted a series of 5 real-time online experiments where pairs of participants were matched to play a sequence of 40 2x2 games together (total N=616). Across the experiments, we varied the exact structure of trials, but all experiments involved trials with reward structures modeled on Prisoner's Dilemma (PD) and Bach or Stravinsky (BoS). We manipulated the presence or absence of a chat interface as a between-pairs manipulation. Despite low levels of use of the communication channel, we found slight differences in outcome between the chat and no-chat groups, where pairs who could communicate were slightly more likely to reach coordinated beneficial outcomes in BoS games. In contrast to the predictions of rational actors, participants in both chat and no-chat conditions generally aligned on both cooperating rather than both defecting in PD. Overall, we were limited in our ability to discern the role of communication channels in game strategies due to low uptake in using the communication channels when available.

@@

In studies of conversation, there's work on how people align with each other in their word use when doing joint task and how repeatedly referring to the same targets in a matching game leads to the formation of linguistic conventions. As reference is a key part of other linguistic speech acts such as negotiation, coordinating actions, or persuasion, one might wonder whether referential coordination can also be seen during interactions in pursuit of these more complex goals. In a similar vein, more complex situations may allow for linguistic alignment and convention formation not just of terms of reference but of terms for entire strategic approaches. One could also imagine alignment in approach, creating a conventional structure for how to coordinate or negotiate.

@@

In prior work, we tried to look at the role of language in a 3-player negotiation game where incentives were either fully-aligned or competitive between players [@mankewitz2021; @boyce2023flowers]. We found some evidence that language use improved performance, and that the referential language used to pick out targets showed some signs of conventionalization over time. However, this was a game we had created, so there were not theoretical or analytic predictions about what the optimal strategies were either with or without language, and therefore, when or how communication (via language) might improve performance.

Game theory provides a formalism for understanding rational action when different actors have different incentive structures over outcomes that are jointly determined by the choices of all the actors. There is some theoretical work analyzing how communication availability may shift the strategies in games and thus when communicative acts are credible signals versus cheap talk [@cooper1992; @foerster2021; @baliga2002]. Game theory economic games have been used to empirically study questions such as social prediction [@vanbaar2021], emotional signals [@demelo2020], and coordination [@duguid2014]. Some studies test predictions about the role of communication in economic games [@crandall2018; @cross2020; @crawford2019a].

@@

In Bach or Stravinsky (Figure \ref{payoff-BoS}), both players only receive rewards if they choose the same option. However, player X prefers the AA option and player Y prefers the BB option. Thus, there are two Nash equilibria, either AA or BB. In a one-shot game of BoS, there is no way for participants to break the symmetry. Communication is predicted to be potentially useful, if it allows participants to break this symmetry and agree on which cell to align on.

@@

## Participants
The number of participants included in each condition (after incomplete games were excluded) is shown in Table \ref{participants}. We recruited participants via Prolific and directed them to our website where they were paired up to play a game with another participant in real time.

## Procedure
All experiments were implemented in Empirica v1 [@almaatouq2021]. The experiment consisted first of consent and instruction.

@@

After the three minutes, participants moved onto the main experiment.  On each trial, they saw a payout grid for differently colored treasure chests and were asked to click which chest they wanted to open (Figure \ref{ui}). Each pair completed 40 trials, with the payoffs dependent on the experiment condition (see below). Participants in the "chat" condition could message their partner freely during the trials using a chat interface; participants in the "no-chat" condition did not have access to the chat during the main experiment.

In experiments 1, 2, 3, and 5, participants saw a running bonus score for themselves and their partner, as their bonus payment was proportional to the sum of the rewards they earned. On experiment 4, participants were instead paid proportional to the results of 4 randomly selected trials, and were not shown a cumulative payoff score.

@@

## Trial types 
We sampled different types of trials across different experiments. For each trial type, we constructed trials by randomly sampling reward values subject to some constraints. Examples of each trial type are shown in Figure \ref{payoff-matrix}.

@@

**Expt 4** (https://osf.io/fymjc): In experiment 4, we aimed for a set-up that was closer to the predictions from game theory. We removed the points counter, given that participants in prior experiments sometimes used the visible money earned so far as a locus for coordination. We switched from providing a bonus proportional to the overall points earned to an incentive compatible bonus that sampled 4 trials at random and paid out those points. Because the games that mattered were unknown and randomly sampled, we reasoned that the rational strategy might be closer to the formal predictions of one-shot games. Trials were sampled from a 50/50 distribution of easy PD and normal BoS.

@@

exit <- exit_1 |>
  bind_rows(exit_2) |>
  bind_rows(exit_3) |>
  bind_rows(exit_4) |>
  bind_rows(exit_5) |>
  select(expt, chat_cond, pct) |>
  mutate(chat_cond = ifelse(chat_cond == "chat", "chat", "no-chat"))

@@

```{r bosquad, fig.cap="For BoS trials, how frequently pairs made coordinated selections that recieved rewards. \\label{bosquad} ", fig.height=3, fig.width=8}
outcome %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond), y = outcome_parity, color=combo_type, group=combo_type)) +
  facet_grid(~ expt) +
  scale_color_manual(values=pal)+
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  coord_cartesian(ylim=c(0,1))+
  labs(y = "Good outcome?", x="Chat condition", color="")+
  theme(legend.position="bottom")
```
For PD trials, there are three distinct outcomes. If both players cooperate, the outcome is AA and each gets a moderately high payoff. If both players defect, the outcome is BB, the Nash equilibrium, and each player gets a low payoff. If one player cooperates and the other defects, the outcome is AB (or BA), and one player gets a high payoff and the other gets 0. From a perspective of wefare-maximization, the selection with best total reward is AA in easy PD and AB/BA in hard PD. We expect normal PD to generally be the of the type easy PD; and again note that the errors in displaying off-diagonal rewards would primarily affect hard PD type trials, where off diagonal options are more attractive. The rates of selecting each of the three outcome options by experiment and conditions are shown in Figure \ref{pdquad}.

@@

Overall, the results based on condition assignment are consistent with the prediction that language would be more beneficial for BoS than for PD. However, condition differences were small and often noisy (presumably due to low uptake of using language), and so the experiments do not provide compelling evidence. Additionally, the analytic import of using well-studied game theory scenarios with predictions about the utility of language is weakened by the inconsistency between predicted behaviors in the PD condition and actual selections.

@@

```{r}
outcome_chatted <- outcome |> 
  mutate(chatted = case_when(
    chat_cond =="nochat" ~ "no-chat",
    is_chat == 1 ~ "chatty",
    is_chat == 0 ~ "quiet"
  ) |> factor(levels=c("chatty", "quiet", "no-chat")))
```

We repeat the quadrant analyses, now with a three-way split between "chatty" games who used the chat on the given trial, "quiet" games who were in the chat condition, but produced no chat messages on the given trial, and "no-chat" games were assigned to the no-chat condition.

```{r, fig.cap="For BoS trials, how frequently pairs made coordinated selections that recieved rewards as a function of whether chat was available or used that trial. \\label{bosquad-uptake} ", fig.height=3, fig.width=8}
outcome_chatted %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = chatted, y = outcome_parity, color=combo_type, group=combo_type)) +
  facet_grid(~ expt) +
    coord_cartesian(ylim=c(0,1))+
  scale_color_manual(values=pal)+
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  labs(y = "Good outcome?", x="Chat usage on trial", color="")+
  theme(legend.position="bottom")
```
For BoS results, shown in Figure \ref{bosquad-uptake}, chatty trials had the best outcomes, with high levels of aligning. Depending on experiment, quiet trials generally patterned with the no-chat condition, except in the first experiment that was entirely BoS. Thus, it appears that there may be slight benefits to chatting that spill over onto other trials, perhaps via setting up conventions for determining what strategy to use to break the symmetry. Strategy set-up may be easiest when the trials are most homogeneous. However, most of the overall slight benefits associated with the chat condition are being driven by the small number of games and trials that used the chat.

@@

We similarly break down the PD results in terms of language use per trial in Figure \ref{pdquad}. Trials that used chat were slightly more consistent at selecting the cooperate option on easy and normal PD (although this in particular could reflect selection effects for careful or attentive participants). The larger difference in selections between trials that used chat and those that did not was on the hard PD trials where selection of the AB/BA option appears to be primarily driven by chat usage.

Overall, the slight differences observed in the chat condition and no-chat conditions appear to primarily come from the relatively small number of trials that actually used the chat while quiet trials generally patterned with behavior from the no-chat condition.

@@

Despite low levels of use of the communication channel, we found slight differences in outcome between the chat and no-chat groups, where pairs who could communicate were slightly more likely to reach coordinated beneficial outcomes in BoS games and slightly more likely to choose welfare-maximizing unequal outcomes in hard PD games. In contrast to the predictions of rational actors, participants in both chat and no-chat conditions generally aligned on both cooperating rather than both defecting in PD. The tendency to cooperate has previously been observed in experiments with repeated PD-style games [@rand2013]. Overall, we were limited in our ability to discern the role of communication channels due to low uptake in using the communication channels when available.

@@

In conclusion, we still think there are interesting questions about how communication, and linguistic communication in particular, might affect what strategies around coordination and competition can be used in strategic games. Coming from the perspective of convention formation in reference games, we are still interested in when and how people can form procedural or "meta" conventions around how to negotiate with each other rather than just around referents. However, over the course of these experiments, we have learned that there are methodological and implementational barriers. We were able to mostly overcome the initial issue with participants thinking they were playing with bots by adding an initial short free chat between each partnership. However, we were not able to overcome other problems. We had difficulties finding tasks that have tractable theoretical predictions but are also sufficiently engaging that humans will take them seriously, and compromised on a design that was not ideal from either perspective. Additionally, while we previously had no problems getting participants to use linguistic communication in online chat-box mediated tasks where the goal was reference [@boyce2024], we had difficulty getting participants to use chat in strategic contexts [both here and in @mankewitz2021].
